---
title: "HAPR probit HRS Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HAPR probit HRS Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(tidyverse)
devtools::load_all()


hrs_data<-read_rds("~/Downloads/Research/hrs_wave.rds")
hrs_data_person<-read_rds("~/Downloads/Research/hrs_person.rds")

hrs_full<-merge(x = hrs_data, y = hrs_data_person, by = "id", all.x = TRUE)

hrs_na_omit <- na.omit(hrs_full |> select(heart_problem,prs_cad,wave,age,bmi,smoking, total_wealth,cholesterol))
```

A polygenic score (PGS) aggregates millions of individual loci across the human genome and weights them by the strength of their association to produce a single quantitative measure of genetic risk. We are using a dataset for HRS respondents who provided salivary DNA between 2006 and 2012. We observe disease status (CAD/heart problem) and a polygenic score. We also have covariates: wave, age, bmi, cholesterol, and wealth. In this data, we are restricting values to those where no covariates have N/A values.

```{r, echo=FALSE}

model <- glm(hrs_na_omit$heart_problem~(hrs_na_omit$prs_cad+hrs_na_omit$wave+hrs_na_omit$age+hrs_na_omit$bmi+hrs_na_omit$cholesterol+hrs_na_omit$total_wealth),family='binomial')

covariates <- hrs_na_omit |> select(wave,age,bmi,smoking, total_wealth,cholesterol)
R2logit<- function(y,model){
    R2<- 1-(model$deviance/model$null.deviance)
    return(R2)
    }
r2_current_value<-R2logit(hrs_na_omit$heart_problem,model)

```

We are running the HAPR function here. We calculated the current $R^2$ based on the pseudo $R^2$ calculated by the logistic model of covariates and polygenic score.The estimated heritability of CAD is around 50%. HAPR combines these heritability estimates with the current correlations to estimate a model of the future polygenic score.

The main estimation function is `hapr()`.

```{r, echo=FALSE}
hapr_fit <- hapr(
  y = hrs_na_omit$heart_problem, # Outcome
  gc = hrs_na_omit$prs_cad, # Polygenic score
  w = covariates, # Covariates
  model_type = "probit",
  r2_current = r2_current_value, # Current R-squared of the polygenic score
  r2_future = 0.5 # Future R-squared of the polygenic score, estimate of heritability
)
print(hapr_fit)
```

That was easy! Behind the scenes, `hapr()` run a feasible probit regression of CAD on the polygenic score and covariates. It then combines this with heritability estimates to estimate the "true model" of how CAD is predicted by the true polygenic score. The reported `beta` is the coefficient of the true polygenic score.

It is intuitive to see what is going on if we simulate the full model with the true polygenic score:
```{r, echo=FALSE}
simulated_data <- hapr_simulate(hapr_fit, covariates)
head(simulated_data)
```

The `simulated_data` contains the true polygenic score, the observed polygenic score, and the outcome. This is the dream dataset we would like to have to evaluate the utility of the polygenic score. Moreover, with the predict function, we can add the predictions from each model to the data.

```{r}
risk_dataset <- predict(hapr_fit, simulated_data)
head(risk_dataset)
```

We can now compare how the current and future polygenic scores add value relative to the traditional risk factors. One simple way are scatterplots of the predicted risk using the polygenic score vs the predicted risk using the traditional risk factors. If the polygenic score adds little extra value, we should see the points clustered around the diagonal.

```{r}
# Plot for current polygenic score
ggplot(risk_dataset, aes(x = y_hat_w, y = y_hat_gc_w)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  coord_fixed(ratio = 1) +  # Ensures 1:1 aspect ratio
  theme_minimal() +
  labs(x = "Predicted risk using traditional factors", 
       y = "Predicted risk using current polygenic score")

# Plot for future polygenic score
ggplot(risk_dataset, aes(x = y_hat_w, y = y_hat_gf_w)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  coord_fixed(ratio = 1) +  # Ensures 1:1 aspect ratio
  theme_minimal() +
  labs(x = "Predicted risk using traditional factors", 
       y = "Predicted risk using future polygenic score")
```
