---
title: "HAPR cox"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HAPR cox}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(tidyverse)
devtools::load_all()

# Set seed for reproducibility
set.seed(123)

# Number of patients
n <- 10000

# Generate gender
gender <- sample(c("male", "female"), n, replace = TRUE)

# Generate prevent_score
prevent_score <- ifelse(
  gender == "male", rbeta(n, 1, 12), rbeta(n, 1, 24)
)

# Calculate prevent_index (inverse Gaussian CDF of prevent_score)
prevent_index <- qnorm(prevent_score)

# Generate V: Gaussian with variance 1/3
V <- rnorm(n, mean = 0, sd = sqrt(1/3))

# Calculate gf: V + prevent_score/10
gf <- V + prevent_score/2

# Calculate gc: V plus Gaussian with variance 1
gc <- V + rnorm(n, mean = 0, sd = 1)

# Normalize gc
gc <- (gc - mean(gc)) / sd(gc)

# Calculate heart_disease_index: prevent_score + 0.42 * gf
heart_disease_index <- prevent_index + V

# Heart disease index
heart_disease_index <- prevent_index + V

# Probability of ASCVD
prob_ascvd <- pnorm(heart_disease_index)

# Simulate event times
relative_risk <- prob_ascvd / 7
event_time <- rexp(n, rate = relative_risk)

# Apply random censoring
censor_time <- runif(n, 0, 10)
status <- as.integer(event_time <= censor_time)
observed_time <- pmin(event_time, censor_time)

# Create dataset
cox_data <- tibble(
  gender = factor(gender),
  prevent_score = prevent_score,
  prevent_index = prevent_index,
  V = V,
  gc = gc,
  heart_disease_index = heart_disease_index,
  prob_ascvd = prob_ascvd,
  time = observed_time,
  ascvd = status
)

# Select relevant variables
heart_disease_data <- cox_data |> 
  select(time, ascvd, polygenic_score = gc, prevent_score, gender)
```

We have electronic health records on potential cardiac patients, as in the introductory probit vignette. We observe disease status (ASCVD) and a polygenic score. We also have covariates: PREVENT score (a standard risk estimate based on traditional risk factors) and gender. Here, we assume that patients are followed by different lengths of time, and that there is censoring, as is common in electronic health records.

```{r, fig.alt = "Plot of Time to Event status", fig.width=6, fig.asp=9/16}
# Display the first few rows of our data
head(heart_disease_data)

# Plot time versus status
ggplot(heart_disease_data, aes(x = time, fill = factor(status))) +
  geom_histogram(binwidth = 1, alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Time to Event vs Event Status",
    x = "Observed Time",
    y = "Event Status (1 = event occurred, 0 = censored)"
  )
```

Fortunately, the two-step estimator behind `hapr` works well with many statistical models, including survival models. Here we use the standard Cox proportional hazards model. The workflow is the same as in the probit:

```{r}
library(hapr)
# Fit the model
y <- survival::Surv(
    heart_disease_data$time,
    heart_disease_data$ascvd) # Outcome

covariates <- heart_disease_data |> 
  select(prevent_score, gender) # Covariates

fit <- hapr(
  y = y, # Outcome
  gc = heart_disease_data$polygenic_score, # Polygenic score
  w = covariates, # Covariates
  model_type = "cox", # Model type
  r2_current = 0.02, # Current R-squared of the polygenic score
  r2_future = 0.2 # Future R-squared of the polygenic score
)

# # Display the model summary
print(fit)
```

With the model fitted and the simulated data, we can evaluate the performance of the current and future polygenic scores. A good place to start is to compare survival curves.

```{r, fig.alt = "Survival Curves", fig.width=6, fig.asp=9/16}
# Plot survival curves
simulated_data <- hapr_simulate(fit, covariates)
survival_curves_gf_w <- hapr_survfit(fit, covariates = "gf_w", newdata = simulated_data)
plot(survival_curves_gf_w)

survival_curves_gc_w <- hapr_survfit(fit, covariates = "gc_w", newdata = simulated_data)
plot(survival_curves_gc_w)

survival_curves_w <- hapr_survfit(fit, covariates = "w", newdata = simulated_data)
plot(survival_curves_w)
```

