---
title: "HAPR probit HRS Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HAPR probit HRS Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(tidyverse)
devtools::load_all()


hrs_data<-read_rds("~/Downloads/Research/hrs_wave.rds")
hrs_data_person<-read_rds("~/Downloads/Research/hrs_person.rds")

hrs_full<-merge(x = hrs_data, y = hrs_data_person, by = "id", all.x = TRUE)

covariates <- hrs_full |> select(wave,age,bmi,smoking, total_wealth,cholesterol)
```

We have electronic health records on potential cardiac patients. The patients are followed for 10 years. We observe disease status (ASCVD) and a polygenic score. We also have covariates: PREVENT score (a standard risk estimate based on traditional risk factors) and gender.

```{r, echo=FALSE}
model <- glm(hrs_full$heart_problem~(hrs_full$prs_cad+hrs_full$wave+hrs_full$age+hrs_full$bmi+hrs_full$cholesterol+hrs_full$total_wealth),family='binomial')
#model <- lm(hrs_full$heart_problem~(hrs_full$wave+hrs_full$age+hrs_full$bmi+hrs_full$cholesterol+hrs_full$total_wealth))
R2logit<- function(y,model){
    R2<- 1-(model$deviance/model$null.deviance)
    return(R2)
    }
r2_current_value<-R2logit(hrs_full$heart_problem,model)

```

We are running the HAPR function here. We calculated the current $R^2$ based on the pseudo $R^2$ calculated by the logistic model of covariates and polygenic score.The estimated heritability of ASCVD is around 50%. HAPR combines these heritability estimates with the current correlations to estimate a model of the future polygenic score.

The main estimation function is `hapr()`.

```{r, echo=FALSE}
hapr_fit <- hapr(
  y = hrs_full$heart_problem, # Outcome
  gc = hrs_full$prs_cad, # Polygenic score
  w = covariates, # Covariates
  model_type = "probit",
  r2_current = r2_current_value, # Current R-squared of the polygenic score
  r2_future = 0.5 # Future R-squared of the polygenic score, estimate of heritability
)
print(hapr_fit)
```

These are the results of the HAPR function. As we can see, there are coefficients of the hapr function.

```{r}
nrow(covariates)
sum(is.na(covariates))
```

We can't run the hapr_simulate on this dataset due to the presence of NA values which will make the HAPR simulate function return the error 
    `Error in hapr_simulate(hapr_fit, covariates): dims [product 139832] do not match the length of object [281208]. `
We will create another document with the NA's removed and simulate it.

