---
title: "HAPR Weibull survival MLE"
output:
  rmarkdown::html_vignette:
    syntax_highlighting: none
vignette: >
  %\VignetteIndexEntry{HAPR Weibull survival MLE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(hapr)
library(tidyverse)

set.seed(123)
```

In this vignette we use a parametric survival model with a Weibull baseline. We create a synthetic heart‑disease dataset with time‑to‑event outcomes and censoring. The outcome is a pair `(event_time, event_status)`, where `event_status = 1` indicates disease onset and `0` indicates censoring.

```{r}
n <- 1000

# Covariates (panel-style data with time to disease)
gender <- sample(c("male", "female"), n, replace = TRUE)
prevent_score <- rbeta(n, 1, 12)

covariates <- tibble(
  prevent_score = prevent_score,
  gender = factor(gender)
)

# Build w in the same format required by hapr()
w <- model.matrix(~ prevent_score + gender, data = covariates)[, -1, drop = FALSE]

# Latent genetic structure
theta <- c(0.0, 0.1, -0.2)  # (Intercept), w1, w2
var_v <- 0.15
var_epsilon <- 0.70

v <- rnorm(n, 0, sqrt(var_v))
e <- rnorm(n, 0, sqrt(var_epsilon))

w_with_int <- cbind(1, w)
gf <- v + w_with_int %*% theta
gc <- gf + e

# Weibull survival model: scale = exp(linpred), shape k = exp(log_k)
beta_g <- 0.6
beta_w <- c(0.1, -0.2, 0.15)  # (Intercept), w1, w2
log_k <- 0.0
k <- exp(log_k)

linpred <- beta_g * gf + w_with_int %*% beta_w
scale <- exp(linpred)
event_time <- scale * (-log(runif(n)))^(1 / k)

# Add independent censoring
censor_time <- rexp(n, rate = 0.2)
event_status <- as.numeric(event_time <= censor_time)
event_time <- pmin(event_time, censor_time)

head(tibble(event_time, event_status, gc, prevent_score, gender))
```

We fit the Weibull survival model using the MLE estimator. The only model‑specific likelihood parameter is the shape `k`, which is optimized as `log_k`.

```{r}
improvement_ratio <- 1 / (1 - var_epsilon)
start_beta <- rep(0, ncol(w) + 2)
start_delta <- c(log_k = 0)

mle_fit <- hapr_mle_survival(
  event_time = event_time,
  event_status = event_status,
  gc = gc,
  w = w,
  improvement_ratio = improvement_ratio,
  model_type = "weibull",
  start_beta = start_beta,
  start_delta = start_delta,
  control = list(maxit = 150)
)

print(mle_fit)
```

Notes:

- `w` must be a numeric matrix with no intercept column. The order must match the `w` used in the initial call.
- `event_status` is numeric `0/1` (or logical). `event_time` must be non‑negative.
- MLE standard errors ignore first‑stage uncertainty, so they are typically smaller than two‑stage standard errors.
